Transitioning towards Continuous Delivery in the B2B Domain: A Case Study

### 2014-12-06 paper.tex
Added content

\newsavebox\mybox
\newenvironment{aquote}[1]
  {\savebox\mybox{#1}\begin{quotation}}
  {\signed{\usebox\mybox}\end{quotation}}
%
\begin{document}
%
\mainmatter              % start of the contribution
%
\title{Transitionin towards Continuous Delivery in the B2B Domain: A Case Study}
%
\titlerunning{Abbreviated title}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Olli Rissanen \and J{\"u}rgen M{\"u}nch
%Jeffrey Dean \and David Grove \and Craig Chambers \and Kim~B.~Bruce \and
}
%
\authorrunning{Olli Rissanen}   % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
\tocauthor{Olli Rissanen}
%
\institute{University of Helsinki, Helsinki, Finland,\\
\email{olli.rissanen@aalto.fi}
}

@@

\begin{abstract}        % give a summary of your study
Delivering more value to the customer is the goal of every software company. In modern software business, delivering value in real-time requires a company to utilize real-time deployment of software to expose features to users faster, and to shorten the feedback loop. This allows for faster reaction times, ensuring the development is focused on features providing real value. This study investigates a practice known as continuous delivery as means of providing value for the customers in real-time. Continuous delivery is a development practice where the software functionality is deployed continuously to customer environment. This process includes automated builds, automated testing and automated deployment. As a part of this paper, a case study is conducted in a medium-sized software company. The research objective is to analyze the challenges, benefits and organizational aspects of continuous delivery in the B2B domain. The data is collected from interviews conducted on members of two teams developing two different software products. The results suggest that technical challenges are only one part of the challenges a company encounters in this transition. For continuous delivery, the company must also address challenges related to the customer and procedures. The core challenges are caused by having multiple customers with diverse environments and unique properties, whose business depends on the software product. Some customers also require to perform manual acceptance testing, which slows down production deployments. The benefits found from continuous delivery support the case company in solving many of its business problems. The company can expose the software functionality to the customers from an earlier stage, and better utilize customer feedback.
%                         please supply keywords within your abstract
\keywords {Continuous delivery, Development process, B2B}
\end{abstract}
%
\section{Introduction}

@@

Continuous delivery is a design practice aiming to shorten the delivery cycles and by automating the deployment process. It is an extension to continuous integration, where the software functionality is deployed frequently to customer environment, and the software is developed in a way that it is always ready for release. While continuous integration defines a process where the work is automatically built, tested and frequently integrated to mainline \cite{duvall2007continuous}, continuous delivery adds automated acceptance testing and automated deployment. Continuous delivery therefore attempts to deliver an idea to users as fast as possible.

This study is an exploratory deductive case study, which explores how continuous delivery be applied in the case company. While existing studies of applying the development models exist \cite{neely2013continuous}, none of the studies focuses specifically in the B2B domain. This study specifically aims to identify the main requirements, problems and key success factors with regards to continuous deliver in the B2B domain. Extending the development process towards continuous delivery requires a deep analysis of the current development and deployment process, seeking the current problems and strengths. Adopting continuous deliveryalso requires understanding the requirements of continuous delivery, and restrictions caused by the developed software products.

In this research, the units under the study are two teams and the two software products developed by these teams. By focusing on two different products, a broader view on the application and consequences of these development approaches can be gained. The first product, a marketing automation called Dialog, is used through an extensive user interface. The second product under inspection is CDM, a Master Data Management \cite{loshin2010master} solution running as an integrated background application. The objective of this study can therefore be summarized as follows:

@@

\noindent To rationalize the decision to adopt continuous delivery in a company, the actual benefits to the business have to be identified. This research question aims to identify clear objectives for what is achieved by adopting continuous delivery. Sections of the interview aim to identify the current perceived problems of the case company related to deployment, product development, collecting feedback and guiding the development process. These problems are then compared to the benefits of the approach found from the literature.
\newline

\subsection{Problem}

\subsection{Background}
Continuous delivery is an extension to continuous integration, where the software functionality is deployed frequently at customer environment. While continuous integration defines a process where the work is automatically built, tested and frequently integrated to mainline \cite{duvall2007continuous}, often multiple times a day, continuous delivery adds automated acceptance testing and deployment to a staging environment. The purpose of continuous delivery is that as the deployment process is automated, it reduces human error, documents required for the build and increases confidence that the build works \cite{cdbook}. Continuous delivery solves the problem of how to deliver an idea to users as quickly as possible.

%release cycles
In an agile process software release is done in periodic intervals \cite{cockburn2002agile}. Compared to waterfall model it introduces multiple releases throughout the development. Continuous delivery, on the other hand, attemps to keep the software ready for release at all times during development process \cite{cdbook}. Instead of stopping the development process and creating a build as in an agile process, the software is continuously deployed to customer environment. This doesn't mean that the development cycles in continuous delivery are shorter, but that the development is done in a way that makes the software always ready for release.
\begin{figure}[h]
  \centering
  \includegraphics[width=5.0in]{rtvd.jpg}
  \caption{Continuous integration, delivery and deployment.}
  \label{fig2}
\end{figure}

%\begin{figure}
% use the \includegraphics command from the graphics package
% (N.B. put \usepackage{graphics} into your preamble then)
% for inclusion of your images. A sample call could read:
% \includegraphics{myimage}
%\vspace{2.5cm}
%\caption{This is the caption of the figure displaying a white eagle and
%a white horse on a snow field}
%\end{figure}
%delivery vs deployment
Continuous delivery differs from continuous deployment. Refer to Fig. \ref{fig2} for a visual representation of differences in continuous integration, delivery and deployment. Both continuous delivery and continuous deployment include automated deployment to a staging environment. Continuous deployment includes deployment to a production environment, while in continuous delivery the deployment to a production environment is done manually. The purpose of continuous delivery is to prove that every build is deployable \cite{cdbook}. While it necessarily doesn't mean that teams release often, keeping the software in a state where a release can be made instantly is often seen beneficial.

\subsection{Structure of the study}

This study is organized as follows. The first chapter is the introduction at hand. The second and third chapter summarize the relevant literature and theories to position the research and to educate the reader on the body of knowledge and where the contributions are intended. The fourth chapter briefly explains the usage of case study and qualitative research as research methods, to provide background for the decision to utilize them in this particular study. The fifth chapter presents the research design: objectives, context and applied methods. The findings are then presented in the sixth chapter, organized according to the research questions. The seventh chapter summarizes and interprets the main results. Finally, the eight chapter summarizes the results of study and answers to the research question, discusses the limitations of the study and introduces further research avenues.
%

\section{Related works}

\bigskip
\noindent \textbf{Research objective}
To analyze the software development process transition towards shorter feedback cycle and data-driven decisions from the perspective of the case company in the B2B domain, with respect to continuous delivery.

\subsection{Research method}
%Methods
The primary source of information in this research are semi-structured interviews \cite{runeson2009guidelines} performed within the CDM and Dialog teams. The interview consists of pre-defined themes focusing on current development process, current deployment process, current interaction with customers, the software products and future ways with continuous delivery. Data is also collected through the product description documents and development process documents to verifying and supplementing the interview data. Data triangulation is implemented by interviewing multiple individuals in different positions of the teams. Methodological triangulation is implemented by collecting documentary data regarding the development process from internal documents, and by including observations made by the author in the analysis.

@@

%Data analysis
In this case study the data analysis is based on template analysis, which is a way of thematically analysing qualitative data \cite{king1998template}. The initial template was first formed by exploring the qualitative data for two themes: development process and deployment of software. Through multiple iterations of the data, multiple subthemes were then added to the two existing themes by further coding the data. Attention was paid to different roles of the interviewees.
\subsection{Execution}

@@

%When to use CD: http://www.continuousagile.com/unblock/cd_costs_benefits.html
%You provide an online service: Web, SaaS, PaaS, online gaming, online big data, or mobile apps with Web service back ends. If you provide online services, you need to move to continuous delivery, or your competitors will race ahead of you.

This section addresses the RQ1 (i.e. What are the B2B specific challenges of continuous delivery?). The challenges regarding continuous delivery are analyzed in three areas: technical, procedural and customer. Technical aspect includes the environmental challenges, configural challenges and other challenges related to the software product and its usage. Procedural aspect includes the challenges regarding the development process of software. Customer aspect consists of the customer interaction process and customer commitment.

@@

\begin{center}
\begin{table}[htb]
    \begin{tabular}{ | p{12cm} |}
    \hline
    \textbf{Specific problem} \\ \hline
    Downtime is critical for certain customers \\ \hline
    Automated testing has to be built on top of a matured software product \\ \hline
    Software is often integrated to multiple third party applications \\ \hline
    Software is often accompanied by multiple external components \\ \hline
    There exists multiple different configurations due to having multiple customers with different specifications \\ \hline
    Transferring the software product to diverse customer-owned environments requires different deployment configurations \\ 
    \hline
    \end{tabular}
    \caption{Technical challenges in continuous delivery.}
    \end{table}
\end{center}
%Downtime
Downtime of the case company's products can be fatal. According to the Dialog product owner, downtime causes end-users being unable to perform their job. Downtime can also interrupt ongoing customer tasks, possibly losing critical data in the progress. With CDM, downtime can cause other integrated systems to shut down due to too many failing requests. Currently the deployment time for both projects is negotiated with the customer to prevent these cases, and the version deployments are done when the system can be closed for a short period of time. In the continuous delivery process the deployment pipeline has to be designed to address the issue of downtime.

@@

\begin{center}
\begin{table}[htb]
    \begin{tabular}{ | p{12cm} |}
    \hline
    \textbf{Specific problem} \\ \hline
    User acceptance testing environment is a requisite for production release \\ \hline
    The development process drifts towards small feature branches from long-lived feature branches \\ \hline
  Triggering the compilation and deployment of a modular project to maintain integrity is hard \\ \hline
  The software has to be deployed to multiple customers \\ \hline
  Versioning is affected by having different customer profiles of the product \\ \hline
  Responsibility of deploying moves towards developers \\ \hline
  Management and sales loses track of versions \\
    \hline
    \end{tabular}
    \caption{Procedural challenges in continuous delivery.}
    \end{table}
\end{center}

@@

\begin{center}
\begin{table}[htb]
    \begin{tabular}{ | p{12cm} |}
    \hline
    \textbf{Specific problem} \\ \hline
    Some customers are reluctant towards new versions \\ \hline
    Customers are trained to use a certain version, and modifications confuse the users \\ \hline
    Changelogs are especially important, since as versions are released faster the customers become less aware on what has changed  \\ \hline
    Pilot customer is required for developing the continuous delivery process \\ \hline
    Acceptance testing is performed by both the company and the customers, and requires a lot of resources from the customers \\ \hline
  Production deployment schedule has to be negotiated with the customer \\ \hline
  Ongoing critical tasks by users cannot be interrupted by downtime \\ 
    \hline
    \end{tabular}
    \caption{Customer challenges in continuous delivery.}
    \end{table}
\end{center}

@@

While the data collection was applied to the teams developing the software products, the upper management of the company were not interviewed. The variety of responses in other teams was not explored, or the multitude of other different possible contexts in software development in the B2B domain. The responses from the interviewees mostly differentiated only based on the software product being developed, and the teams were quite similar in terms of composition and experience.
\subsection{Future research}
The individual aspects where challenges were found could be more deeply analyzed for both continuous delivery and continuous experimentation. Also, grouping the challenges into different groupings can yield interesting results. In the case of continuous experimentation, researching dependencies between problems in different groups could yield interesting results by providing a detailed view on the core challenges.

An interesting research question for further research is also how the core findings of the study can be transferred to other companies working in the B2B domain with different software products. This includes investigating deviations found in the challenges and benefits presented in this study. The challenges found in the transition phases also have to be validated by adopting the respective development model in practice.

>>>>

\newsavebox\mybox
\newenvironment{aquote}[1]
  {\savebox\mybox{#1}\begin{quotation}}
  {\signed{\usebox\mybox}\end{quotation}}
%
\begin{document}
%
\mainmatter              % start of the contribution
%
\title{Transitioning towards Continuous Delivery in the B2B Domain: A Case Study}
%
%TODO: check this
%\titlerunning{Abbreviated title}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Olli Rissanen \inst{1,2}, J{\"u}rgen M{\"u}nch \inst{1}
%Jeffrey Dean \and David Grove \and Craig Chambers \and Kim~B.~Bruce \and
}
%TODO: check this
%\authorrunning{Olli Rissanen}   % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
\tocauthor{Olli Rissanen}
%
\institute{Department of Computer Science, University of Helsinki, \\
P.O. Box 68, FI-00014 University of Helsinki, Finland
\and
Steeri Oy, Tammasaarenkatu 5, 00180 Helsinki, Finland
\email{olli.rissanen@aalto.fi, juergen.muench@cs.helsinki.fi}
}

@@

\begin{abstract}        % give a summary of your study
Delivering value to customers in real-time requires a company to utilize real-time deployment of software to expose features to users faster, and to shorten the feedback loop. This allows for faster reaction times, ensuring the development is focused on features providing real value. Continuous delivery is a development practice where the software functionality is deployed continuously to customer environment. Although this practice has been established in some domains such as B2C mobile software, the B2B domain imposes specific challenges. This article imposes a case study that is conducted in a medium-sized software company in the B2B domain. The results suggest that technical challenges are only one part of the challenges a company encounters in this transition. For continuous delivery, the company must also address challenges related to the customer and procedures. The core challenges are caused by having multiple customers with diverse environments and unique properties, whose business depends on the software product. Some customers also require to perform manual acceptance testing, which slows down production deployments. The benefits found from continuous delivery support the case company in solving many of its business problems. The company can expose the software functionality to the customers from an earlier stage, and better utilize customer feedback.
%                         please supply keywords within your abstract
\keywords {Continuous Delivery, Continuous Deployment, Development Process, B2B, Case Study}
\end{abstract}
%
\section{Introduction}

@@

Continuous delivery is a design practice aiming to shorten the delivery cycles and by automating the deployment process. It is an extension to continuous integration, where the software functionality is deployed frequently to customer environment, and the software is developed in a way that it is always ready for release. While continuous integration defines a process where the work is automatically built, tested and frequently integrated to mainline \cite{duvall2007continuous}, continuous delivery adds automated acceptance testing and automated deployment.

This study is an exploratory deductive case study, which explores how continuous delivery can be applied in the case company. While existing studies of applying the development models exist \cite{neely2013continuous}, none of the studies focuses specifically in the B2B domain. This study specifically aims to identify the main requirements, problems and key success factors with regards to continuous deliver in the B2B domain. Extending the development process towards continuous delivery requires a deep analysis of the current development and deployment process, seeking the current problems and strengths. Adopting continuous delivery also requires understanding the requirements of continuous delivery, and restrictions caused by the developed software products.

This study is organized as follows. The first chapter is the introduction at hand. The second chapter summarizes the relevant literature and theories to position the research and to educate the reader on the body of knowledge and where the contributions are intended. The third chapter presents the research design: objectives, context and applied methods. The findings are then presented in the fourth chapter, organized according to the research questions. The fifth chapter interprets the main results. Finally, the sixth chapter summarizes the results of study and answers to the research question, discusses the limitations of the study and introduces further research avenues.

\subsection{Background and Related Work}
Continuous delivery is an extension to continuous integration, where the software functionality is deployed frequently at customer environment. While continuous integration defines a process where the work is automatically built, tested and frequently integrated to mainline \cite{duvall2007continuous}, often multiple times a day, continuous delivery adds automated acceptance testing and deployment to a staging environment. The purpose of continuous delivery is that as the deployment process is automated, it reduces human error, documents required for the build and increases confidence that the build works \cite{cdbook}. Continuous delivery solves the problem of how to deliver an idea to users as quickly as possible.

%release cycles
In an agile process software release is done in periodic intervals \cite{cockburn2002agile}. Compared to waterfall model it introduces multiple releases throughout the development. Continuous delivery, on the other hand, attemps to keep the software ready for release at all times during development process \cite{cdbook}. Instead of stopping the development process and creating a build as in an agile process, the software is continuously deployed to customer environment. This doesn't mean that the development cycles in continuous delivery are shorter, but that the development is done in a way that makes the software always ready for release.

%\begin{figure}
% use the \includegraphics command from the graphics package
% (N.B. put \usepackage{graphics} into your preamble then)
% for inclusion of your images. A sample call could read:
% \includegraphics{myimage}
%\vspace{2.5cm}
%\caption{This is the caption of the figure displaying a white eagle and
%a white horse on a snow field}
%\end{figure}
%delivery vs deployment
Continuous delivery differs from continuous deployment. Both continuous delivery and continuous deployment include automated deployment to a staging environment. Continuous deployment includes deployment to a production environment, while in continuous delivery the deployment to a production environment is done manually. The purpose of continuous delivery is to prove that every build is deployable \cite{cdbook}. While it necessarily doesn't mean that teams release often, keeping the software in a state where a release can be made instantly is often seen beneficial.

@@

\noindent To rationalize the decision to adopt continuous delivery in a company, the actual benefits to the business have to be identified. This research question aims to identify clear objectives for what is achieved by adopting continuous delivery. Sections of the interview aim to identify the current perceived problems of the case company related to deployment, product development, collecting feedback and guiding the development process. These problems are then compared to the benefits of the approach found from the literature.

\subsection{Research design}
In this research, the units under the study are two teams and the two software products developed by these teams. By focusing on two different products, a broader view on the application and consequences of these development approaches can be gained. The first product, a marketing automation called Dialog, is used through an extensive user interface. The second product under inspection is CDM, a Master Data Management \cite{loshin2010master} solution running as an integrated background application. The objective of this study can therefore be summarized as follows:

%Methods
The primary source of information in this research are semi-structured interviews \cite{runeson2009guidelines} performed within the CDM and Dialog teams. The interview consists of pre-defined themes focusing on current development process, current deployment process, current interaction with customers, the software products and future ways with continuous delivery. Data is also collected through the product description documents and development process documents to verifying and supplementing the interview data. Data triangulation is implemented by interviewing multiple individuals in different positions of the teams. Methodological triangulation is implemented by collecting documentary data regarding the development process from internal documents, and by including observations made by the author in the analysis.

@@

%Data analysis
In this case study the data analysis is based on template analysis, which is a way of thematically analysing qualitative data \cite{king1998template}. The initial template was first formed by exploring the qualitative data for two themes: development process and deployment of software. Through multiple iterations of the data, multiple subthemes were then added to the two existing themes by further coding the data. Attention was paid to different roles of the interviewees.

@@

%When to use CD: http://www.continuousagile.com/unblock/cd_costs_benefits.html
%You provide an online service: Web, SaaS, PaaS, online gaming, online big data, or mobile apps with Web service back ends. If you provide online services, you need to move to continuous delivery, or your competitors will race ahead of you.
The challenges regarding continuous delivery are analyzed in three areas: technical, procedural and customer. Technical aspect includes the environmental challenges, configural challenges and other challenges related to the software product and its usage. Procedural aspect includes the challenges regarding the development process of software. Customer aspect consists of the customer interaction process and customer commitment.

@@

%\begin{center}
%\begin{table}[htb]
%    \begin{tabular}{ | p{12cm} |}
%    \hline
%    \textbf{Specific problem} \\ \hline
%    Downtime is critical for certain customers \\ \hline
%    Automated testing has to be built on top of a matured software product \\ \hline
%    Software is often integrated to multiple third party applications \\ \hline
%    Software is often accompanied by multiple external components \\ \hline
%    There exists multiple different configurations due to having multiple customers with different specifications \\ \hline
%    Transferring the software product to diverse customer-owned environments requires different deployment configurations \\ 
%    \hline
%    \end{tabular}
%    \caption{Technical challenges in continuous delivery.}
%    \end{table}
%\end{center}
%Downtime
Downtime of the case company's products can be fatal. According to the Dialog product owner, downtime causes end-users being unable to perform their job. Downtime can also interrupt ongoing customer tasks, possibly losing critical data in the progress. With CDM, downtime can cause other integrated systems to shut down due to too many failing requests. Currently the deployment time for both projects is negotiated with the customer to prevent these cases, and the version deployments are done when the system can be closed for a short period of time. In the continuous delivery process the deployment pipeline has to be designed to address the issue of downtime.

@@

%\begin{center}
%\begin{table}[htb]
%    \begin{tabular}{ | p{12cm} |}
%    \hline
%    \textbf{Specific problem} \\ \hline
%    User acceptance testing environment is a requisite for production release \\ \hline
%    The development process drifts towards small feature branches from long-lived feature branches \\ \hline
%  Triggering the compilation and deployment of a modular project to maintain integrity is hard \\ \hline
%  The software has to be deployed to multiple customers \\ \hline
%  Versioning is affected by having different customer profiles of the product \\ \hline
%  Responsibility of deploying moves towards developers \\ \hline
%  Management and sales loses track of versions \\
%    \hline
%    \end{tabular}
%    \caption{Procedural challenges in continuous delivery.}
%    \end{table}
%\end{center}

@@

%\begin{center}
%\begin{table}[htb]
%    \begin{tabular}{ | p{12cm} |}
%    \hline
%    \textbf{Specific problem} \\ \hline
%    Some customers are reluctant towards new versions \\ \hline
%    Customers are trained to use a certain version, and modifications confuse the users \\ \hline
%    Changelogs are especially important, since as versions are released faster the customers become less aware on what has changed  \\ \hline
%    Pilot customer is required for developing the continuous delivery process \\ \hline
%    Acceptance testing is performed by both the company and the customers, and requires a lot of resources from the customers \\ \hline
%  Production deployment schedule has to be negotiated with the customer \\ \hline
%  Ongoing critical tasks by users cannot be interrupted by downtime \\ 
%    \hline
%    \end{tabular}
%    \caption{Customer challenges in continuous delivery.}
%    \end{table}
%\end{center}

@@

While the data collection was applied to the teams developing the software products, the upper management of the company were not interviewed. The variety of responses in other teams was not explored, or the multitude of other different possible contexts in software development in the B2B domain. The responses from the interviewees mostly differentiated only based on the software product being developed, and the teams were quite similar in terms of composition and experience.
%\subsection{Future research}
%The individual aspects where challenges were found could be more deeply analyzed for both continuous delivery and continuous experimentation. Also, grouping the challenges into different groupings can yield interesting results. In the case of continuous experimentation, researching dependencies between problems in different groups could yield interesting results by providing a detailed view on the core challenges.

%An interesting research question for further research is also how the core findings of the study can be transferred to other companies working in the B2B domain with different software products. This includes investigating deviations found in the challenges and benefits presented in this study. The %challenges found in the transition phases also have to be validated by adopting the respective development model in practice.
